
The following instructions extend the RISC-V Vector (``V'') extension.
We use the following notation to refer to vector registers:
{\tt vrd}  is a vector        destination register,
{\tt vrsX} is a vector source             register
and
{\tt vrt}  is a vector source/destination register.

The following definitions are taken from the draft
vector specification {\tt v0.9}\footnote{\url{https://github.com/riscv/riscv-v-spec/releases/tag/0.9}}.
They are re-produced here to make the following sections easier to
read using only this document.

\begin{itemize}
\item[\ELEN] - {\em Maximum} size of a single vector element in bits.
              $\ELEN \ge 8$ and must be a power of $2$.
\item[\VLEN] - The width in bits of each vector register.
              $\VLEN \ge \ELEN$ and must be a power of $2$.
\item[\SEW]  - The Selected Element Width in bits.
              Each vector register is viewed as $N=\VLEN/\SEW$ elements.
\item[\LMUL] - The Vector Length Multiplier.
              $\LMUL\in\{{1\over 8},{1\over 4},{1\over 2},1,2,4,8\}$.
              When $\LMUL\ge1$, multiple physical vector registers are
              concatenated together to form a single {\em logical} register.
              $\LMUL<1$ is not salient to the vector crypto instructions.
              When $\LMUL>1$, only vector register addresses which are
              integer multiples of $\LMUL$ may be accessed.
\item[\VLMAX]- Is the maximum number of elements a single vector register
              can contain: $\VLMAX=\LMUL*\VLEN/\SEW$.
\item[\EEW]  - Effective Element Width. This is defined per vector register
              operand, and allows for instructions which operate on
              two sets of elements of different widths. E.g. widening
              instructions.
\item[\EMUL] - Effective $\LMUL$. Like $\EEW$, this is defined per vector
              register operand, allowing for widening and narrowing
              operations.
\end{itemize}

The base vector extension has the constraint $\VLEN \ge \ELEN$.
The vector crypto instructions require that $\ELEN \ge 128$ for all
of its instructions, and upto $1024$ for some.
Note that the vector crypto extension {\em does not} require these
large \ELEN values to be supported for all instructions, only those
which require them in order to function.
Each instruction described below includes the minimum \ELEN it
requires to be supported.

% ============================================================================

\import{./}{sec-vector-aes.tex}
\import{./}{sec-vector-clmul.tex}
\import{./}{sec-vector-sha2.tex}
\import{./}{sec-vector-rotate.tex}
\import{./}{sec-vector-grev.tex}

% ============================================================================
