
\begin{itemize}
\item[17/12/19] \begin{itemize}
    \item Initial first version shown to Crypto Task group.
\end{itemize}
\item[10/01/20] \begin{itemize}
    \item Don't recommend funnel shifts for RV64, 128-bit rotations
          are not common enough to justify.
    \item Change LUT4 instruction. Go for very efficient RV64 instruction
          and split 4-to-2 32-bit instruction.
          Avoids extra ternary instructions.
\end{itemize}
\item[20/01/20] \begin{itemize}
    \item LUT4 specification updates for 32-bit variant based on TG discussion.
    \item Initial performance code size impact for SHA256/512 instructions.
\end{itemize}
\item[24/01/20] \begin{itemize}
    \item Split AES proposals into 3 variants.
\end{itemize}
\item[19/02/20] \begin{itemize}
    \item Revisions to AES proposals and added benchmarking results.
    \item Added SHA256/512 benchmark results.
    \item Added AES        benchmark results.
    \item Codified scalar extension design policies.
\end{itemize}
\item[09/03/20] \begin{itemize}
    \item Selected Markku's lightweight AES proposal for RV32.
    \item Added a dedicated RV64 scalar AES proposal.
\end{itemize}
\item[26/03/20] \begin{itemize}
    \item Modified AES RV64 proposal with Barry's improved KeySchedule.
    \item Add Markku's lightweight SM4 proposal.
    \item Add scalar SM3 instructions, similar to SHA2 instructions.
\end{itemize}
\item[17/04/20] \begin{itemize}
    \item Added RV32 instructions for accelerating SHA-512.
\end{itemize}
\item[22/04/20] \begin{itemize}
    \item Update RV64 AES instructions: remove need for hi/lo variants of
          the round instructions.
\end{itemize}
\item[28/04/20] \begin{itemize}
    \item Fix SM4 instruction specs. 
\end{itemize}
\item[08/05/20] \begin{itemize}
    \item Removed Indexed load/store. They weren't used in any of the
          benchmarked workloads, and can be optimised/coded around.
    \item Removed old fused multiply add/accumulate. Not used, ternary.
    \item Removed cmov instruction. Single ternary op not justified since
          it doesn't appear in performance critical places.
\end{itemize}
\item[23/06/20] \begin{itemize}
    \item Transcribed vector instruction descriptions from Richard's
          slide-show.
    \item Changed ``feature discovery'' section to
          ``implementation profiles'' section.
\end{itemize}
\item[16/07/20] \begin{itemize}
    \item Added vector carry-less multiply instructions.
    \item Entropy source proposal all merged in.
\end{itemize}
\end{itemize}
