
% ============================================================================

\subsection{LUT4 Instruction}

\begin{cryptoisa}
lut4lo  rd, rs1, rs2    // RV32, RV64
    for i = 0..7 : rd.4[i] = rs2.4[rs1.4[i]&0x7] if (rs1.4[i] <  8) else 0

lut4hi  rd, rs1, rs2    // RV32, RV64
    for i = 0..7 : rd.4[i] = rs2.4[rs1.4[i]&0x7] if (rs1.4[i] >= 8) else 0

lut4    rd, rs1, rs2    //       RV64
    for i = 0..15 : rd.4[i] = rs2.4[rs1.4[i]]
\end{cryptoisa}

The \mnemonic{lut4*} instructions are used to implement 4-bit lookup tables
on every nibble in a source word.
Many lightweight block ciphers use 4x4 SBoxes:
PRINCE\cite{block:prince},
PRESENT\cite{block:present},
Rectangle\cite{block:rectangle},
GIFT\cite{block:gift},
Twine\cite{block:twine},
Skinny, MANTIS\cite{block:skinny},
Midori \cite{block:midori}.

On RV32, the lookup step is split into two stages.
The \mnemonic{lut4lo} instruction
updates nibbles in the destination with the looked-up value
iff the index is less than eight.
The \mnemonic{lut4hi} version does
the same for index values greater than/equal to eight.
The results are then or'd together.
An example implementation of an $8$-nibble parallel SBox using these
instructions is found in \figref{example:lut4:1}.

On RV64, the entire set of LUT elements fits in a single source register.
The RV64 only \mnemonic{lut4} instruction stores the entire lut in
\rstwo, and uses each nibble in \rsone as an index into it.

Equivalent C code listings for the instructions are found in
\figref{equiv:c:lut4} of Appendix \ref{sec:ccode:lut4}.
Together, the instructions occupy $3$ encoding points.

\begin{figure}[h]
\begin{lstlisting}[style=ASM]
sbox_4bit:
    lut4lo  a3, a0, a1      // a0 = indexes, a1 = low  8 LUT nibbles
    lut4hi  a4, a0, a2      // a0 = indexes, a2 = high 8 LUT nibbles
    or      a0, a3, a4      // Or results together.
    ret                     // Function Return
\end{lstlisting}
\caption{
    Implement 8 parallel 4-to-4 bit SBox operations on RV32 using
    the \mnemonic{lut4hi} and \mnemonic{lut4lo} instructions.
    The inputs to the SBox are stored in \xreg{a0}.
    The high and low $8$ elements of the
    LUT are stored in \xreg{a2} and \xreg{a1} respectively.
}
\label{fig:example:lut4:1}
\end{figure}

