
Presence of the cryptography extension in any form is indicated
by bit $10$ of the {\tt MISA} CSR.
i.e. bit ``K'', because ``C'' is taken and ``K'' is for Krypto.
Table \ref{tab:profiles} describes the implementation profiles for the
Cryptography extension.
Some profiles are orthogonal to one-another, and some overlap.
Where profiles overlap, implementations with both profiles must
implement the union of both feature-sets.

\note{
All of these are yet to be finalised.
The exact mechanism for exposing the Cryptography extension features to
software will likely be defined by the {\tt tech-config} working
group.
}

\begin{table}[h]
\centering
\begin{tabular}{r|lll|lll}
\multirow{2}{*}{Profiles:} &
\multicolumn{3}{c|}{Scalar Profiles} &
\multicolumn{3}{c}{Vector Profiles} \\
                      & \#1  & \#2  & \#3 & \#1  & \#2  & \#3 \\
\hline
PollEntropy            &   x  &   x  &  x  &   x  &  x   &  x  \\
Bitmanip Instructions &   x  &   x  &  x  &      &      &     \\
Scalar SHA2           &   x  &      &  x  &      &      &     \\
Scalar AES            &   x  &      &  x  &      &      &     \\
Scalar SM3            &      &   x  &  x  &      &      &     \\
Scalar SM4            &      &   x  &  x  &      &      &     \\
\hline
Vector AES Per-Round  &      &      &     &   x  &      &  x  \\
Vector AES All-Round  &      &      &     &      &  x   &  x  \\
Vector SHA2 Per-Round &      &      &     &   x  &      &  x  \\
Vector SHA2 All-Round &      &      &     &      &  x   &  x  
\end{tabular}
\caption{
Implementation profiles for the Cryptography extension.
}
\label{tab:profiles}
\end{table}

The scalar profiles all implement the PollEntropy feature and
shared Bitmanip instructions.
Profile $1$ includes the scalar AES/SHA2 extensions, while
Profile $2$ includes the scalar SM3/SM4  extensions.
These were chosen as the most likely to be mutually exclusive in final
products.
Profile $3$ implements all of the scalar features.

The vector profiles are split into the ``per-round'' instructions
(Profile 1),
the ``all-round'' instructions (Profile 2),
or both (Profile 3).
All vector profiles implement the PollEntropy feature.
